Implementation and comparison of register allocation to translate \lstinline!LLVM--! to x86

### 2023-11-24 article.tex
Looks good -Jane

@@

These so-called \textit{branch instructions} that alter the flow of execution are further c

@@

Liveness analysis is the process of finding the program points at which a variable is live. This serves the function of determining which variables are live simultaneously, i.e. in interference/conflict with one another, so cannot be assigned the same physical register without overwriting the value of one another in execution. These can be derived recursively by iteratively applying so-called \textit{dataflow equations} to each node in a \textit{control-flow graph} (CFG) until a stable state/fixed point is reached.

The algorithm described in \cite{AppelAndrewW} and \cite{alma99122774374105763} is based on the following equations for the live-in and live-out variables respectively:
\begin{align}
  in\left[n\right] &= use\left[n\right] \cup (out\left[n\right] - def\left[n\right])\\
  out\left[n\right] &= \bigcup_{s\in succ\left[n\right]} in\left[s\right]
\end{align}
where \(use\left[n\right]\) is the set of all dependent variables, \(def\left[n\right]\) is the set of all variables defines and \(succ\left[n\right]\) is the set of all immediate successor nodes of \(n\).

The structure of the CFG is inherently vague as both  \cite{AppelAndrewW} and \cite{alma99122774374105763}, while based on the same unde

rlying conceptions of dataflow, have different approaches to nodes of a CFG, with the former taking each individual instruction into consideration and the latter each basic block. Either of these approaches are sound

@@

>>>>

% mention basic block early on, they're a contiguous sequence of instructions that do not transfer control flow

@@

%These so-called \textit{branch instructions} that alter the flow of execution are further c

@@

Liveness analysis is the process of finding the program points at which a variable is live. This serves the function of determining which variables are live simultaneously, i.e. in interference/conflict with one another, which cannot be assigned the same physical register without overwriting the value of one another in execution. These can be derived recursively by iteratively applying so-called \textit{dataflow equations} to each node in a \textit{control-flow graph} (CFG) until a stable state/fixed point is reached.

The algorithm described in \cite{AppelAndrewW} and \cite{alma99122774374105763} is based on the following equations for the live-in and live-out variables respectively:
\begin{align}
  in\left[n\right] &= use\left[n\right] \cup (out\left[n\right] - \mathit{def}\left[n\right])\\
  out\left[n\right] &= \bigcup_{s\in \mathit{succ}\left[n\right]} in\left[s\right]
\end{align}
where \(use\left[n\right]\) is the set of all dependent variables, \(def\left[n\right]\) is the set of all variables defines and \(succ\left[n\right]\) is the set of all immediate successor nodes of \(n\).

The structure of the CFG is inherently vague as both  \textit{Modern Compiler Implementation in ML} \cite{AppelAndrewW} and \textit{Compilers : Principles, Techniques, and Tools} \cite{alma99122774374105763}, while based on the same underlying % TODO: fix references
conceptions of dataflow, have different approaches to nodes of a CFG, with the former taking each individual instruction into consideration and the latter each basic block. Either of these approaches are sound

@@

Translating each IR instruction to x86 correctly is a matter of eliminating unintended side-effects. Each LLVM-- instruction is defined to have only one purpose, %% FIXME is purpose the right word?
as concepts such as calling conventions, stack frames, or a \lstinline!FLAGS! register are completely abstracted over in order to remain platform independent.

In contrast, the x86 \textit{instruction set architecture} (ISA) is targeting a  \textit{complex instruction set computer} (CISC) family of processors, the instructions of which perform a much broader set of operations \cite[p.~190]{AppelAndrewW}. This is is in part due to pipelining, i.e. an abstraction over the concrete implementation of the actual processor, which in turn is made for the sake of performance.

An example of this would be the division/remainder operation: since integer division is a non-trivial iterative process wherein both the quotient and remainder is needed throughout, the result of both of these is stored in the \lstinline!%rax! and \lstinline!%rdx! registers respectively. This means two operations are performed simultaneously regardless of which value is used, hence they need to be restored before executing the next instruction, as any variables assigned to \lstinline!%rax! or \lstinline!%rdx! will be overwritten.
